---
title: "Analysis of Gapminder Data"
author: "Tara Franey and Marzia Tamanna"
date: "February 28, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width=8, fig.height=6, fig.path='Figs/', warning=FALSE, message=FALSE)
```

## Gapminder Data

Gapminder is a database of statistics on the wealth and health of nations of five continents.

```{r gapminder, echo=FALSE}
library(curl)
library(ggplot2)
library(dplyr)
library(viridis)
library(knitr)
library(gridExtra)
gapminder_location<-curl(url = "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
gapminder<-read.csv(gapminder_location, stringsAsFactors = FALSE)
gapminderlifemean <- gapminder %>% group_by(continent) %>% summarize(lifeExPmean = mean(lifeExp)) 
gapminderlifese <- gapminder %>% group_by(continent) %>% summarize(lifeExpSE = sd(lifeExp)/sqrt(length(lifeExp)))
gapminderlifeexpdata<-full_join(gapminderlifemean,gapminderlifese)
```

```{r table, echo=FALSE, results='asis'}
kable(summary(gapminder))
```

#Life Expectancy Over Time
By combining these variables and graphing them, we can see relationships. For example, we can look at the mean and standard deviation of life expectancy by continent. 

```{r plot2, echo=FALSE}
ggplot(data=gapminderlifeexpdata, aes(x=continent, y=lifeExPmean, color, fill=continent)) + scale_fill_viridis(discrete=TRUE) + geom_bar(stat = "identity", position = position_dodge()) +   geom_errorbar(aes(ymin=lifeExPmean-lifeExpSE, ymax=lifeExPmean+lifeExpSE), width=.2, position=position_dodge(.9))
```

Variation in life expectancy can be explained by continent, as calculated by a pairwise t-test. 

```{r t-test, include=TRUE}
pairwise.t.test(gapminder$lifeExp, gapminder$continent)
```

We can also look at trends in life expectancy by year, where each dot represents a specific country 

```{r plot1, echo=FALSE}
ggplot(data=gapminder, aes(x=year,y=lifeExp))+geom_point()+ geom_smooth(method=lm) + scale_x_continuous(name="Year") + scale_y_continuous(name="Life Expectancy") + labs(title ="Life Expectancy Over All Countries and Years")
```

```{r model1, include=TRUE}
model1<-lm(lifeExp ~ year, data=gapminder)
```

These data show a linear relationship; however, the linear model graphed above has an R-squared value of `r summary(model1)$r.squared`, possibly due to very high variability in life expectancy in each year. 

Dividing the data in this graph by continent provides additional information. 

```{r plot3, echo=FALSE}
ggplot(data=gapminder, aes(x=year,y=lifeExp, by=country, color=continent))+geom_point() + geom_line()+ scale_x_continuous(name="Year") + scale_color_viridis(discrete=TRUE) + scale_y_continuous(name="Life Expectancy") + labs(title ="Life Expectancy Over All Years Separated by Continent")
```

```{r continentp, echo=FALSE}
modelAsia<-lm(lifeExp ~ year, continent=="Asia", data=gapminder)
modelAfrica<-lm(lifeExp ~ year, continent=="Africa", data=gapminder)
modelEurope<-lm(lifeExp ~ year, continent=="Europe", data=gapminder)
modelAmericas<-lm(lifeExp ~ year, continent=="Americas", data=gapminder)
modelOceania<-lm(lifeExp ~ year, continent=="Oceania", data=gapminder)
```

Another notable visual feature of connecting the points by country is that two clear outliers pop out in terms of sharp life expectancy drops -- one in 1977, one in 1992.A simple code returns the countries these occurred in:

```{r genocide, echo=FALSE}
gapminder %>% filter(year==1977) %>% filter(lifeExp==min(lifeExp)) %>% kable
gapminder %>% filter(year==1992) %>% filter(lifeExp==min(lifeExp)) %>% kable
```

The Khmer Rouge genocide in Cambodia occurred from 1975 to 1979. The Rwandan civil war occurred from 1990 to 1994. 

The r-squared values for the relationship between life expectancy and year per continent using a linear model are much greater: Asia - `r summary(modelAsia)$r.squared`; Africa - `r summary(modelAfrica)$r.squared`; Americas - `r summary(modelAmericas)$r.squared`; Europe - `r summary(modelEurope)$r.squared`; Oceania - `r summary(modelOceania)$r.squared`. 



## Comparison of Mean gdp across the continents from 1952 to 2007

 
```{r continentMeangdp, echo=FALSE}
gapminder %>% group_by(continent) %>% summarise(Meangdp = mean(gdpPercap)) %>%
  ggplot(aes(x=continent,y=Meangdp, color, fill=continent))+ 
  scale_fill_viridis(discrete=TRUE) + geom_bar(stat = "identity", position = position_dodge()) + labs(title ="Mean gdp across the Continents")
```
From this plot we can state that, Africa has the least gdp and Oceania has the highest gdp among the five continents.