---
title: "Gapminder Data"
author: "Tara Franey"
date: "February 28, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width=5, fig.height=5, warning=FALSE, message=FALSE)
```

## Gapminder Data

Gapminder is a database of statistics on the wealth and health of nations.

```{r gapminder, include=TRUE}
library(curl)
library(ggplot2)
library(dplyr)
library(viridis)
gapminder_location<-curl(url = "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
gapminder<-read.csv(gapminder_location, stringsAsFactors = FALSE)
gapminder<-cbind(gapminder,gapminder$pop/1000000)
gapmindermean <- gapminder %>% group_by(continent) %>% summarize(lifeExPmean = mean(lifeExp)) 
gapminderse <- gapminder %>% group_by(continent) %>% summarize(lifeExpSE = sd(lifeExp)/sqrt(length(lifeExp)))
gapminderlifeexpdata<-full_join(gapmindermean,gapminderse)
summary(gapminder)
```

By combining these variables and graphing them, we can see relationships. For example, we can look at the mean and standard deviation of life expectancy by continent. 

```{r plot2, echo=FALSE}
ggplot(data=gapminderlifeexpdata, aes(x=continent, y=lifeExPmean, color, fill=continent)) + scale_fill_viridis(discrete=TRUE) + geom_bar(stat = "identity", position = position_dodge()) +   geom_errorbar(aes(ymin=lifeExPmean-lifeExpSE, ymax=lifeExPmean+lifeExpSE), width=.2, position=position_dodge(.9))
```

in a specific year (each individual country is represented multiple times). Variation in life expectancy can be explained by continent. 

```{r ttest1}
pairwise.t.test(gapminder$lifeExp, gapminder$continent)  
```

We can also look at trends in life expectancy by year, where each dot represents a specific country 

```{r plot1, echo=FALSE}
ggplot(data=gapminder, aes(x=year,y=lifeExp))+geom_point()+ geom_smooth(method=lm) + scale_x_continuous(name="Life Expectancy") + scale_y_continuous(name="GDP per capita") + labs(title ="GDP per capita vs. Life Expectancy")
```

```{r model1, include=TRUE}
model1<-lm(lifeExp ~ year, data=gapminder)
```

These data show a linear relationship; however, the linear model graphed above has an R-squared value of `r summary(model1)$r.squared`, possibly due to very high variability in life expectancy in each year. 

Dividing these points out by 

Dividing the 
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

This is the mean speed of all cars: `r mean(cars$speed)`